---
title: 'DE analysis'
date: '`r Sys.Date()`'
output:
  html_document:
    df_print: kable
    theme: cerulean
  pdf_document:
  github_document:
    toc: true 
    toc_depth: 3 
---


# Libraries

```{r, include=FALSE, message=FALSE}
#library(cacoa)


library(conos)
library(sccore)
library(ggplot2)
library(dplyr)
library(cowplot)
library(viridis) 

library(devtools)
library(Matrix)
library(dplyr)
library(magrittr)
library(org.Hs.eg.db)
library('egg')  #ggarrange

for (n in names(cao$data.object$samples)) {
  cao$data.object$samples[[n]] %<>% pagoda2::Pagoda2$new()
}

library(knitr)
library(optparse)

```



# Setup
```{r}

# ------ Cacoa object ------


# calc.mode <- 'de'
# calc.mode <- 'go'
# calc.mode <- 'gsea'
# calc.mode <- 'figures'

methods.cmp.file <- paste('/home/aigolkina/coda/cacoa-examples/report/report-de/',
                             'log.txt', sep = '')

output.cacoa.file <- paste('/home/aigolkina/coda/cacoa-examples/report/report-de/',
                             dataset.name, '.output.cacoa.rds', sep = '')

output.df.file <- paste('/home/aigolkina/coda/cacoa-examples/report/report-de/',
                             dataset.name, '_resampling_of_samples.rds', sep = '')

path.figures <- '/home/aigolkina/coda/cacoa-examples/report/figures_stab/'

# ------ Methods ------
possible.tests <- c('DESeq2.Wald', 'DESeq2.LRT', 'edgeR',
                   # 'Wilcoxon.edgeR', 't-test.edgeR',
                   'Wilcoxon.DESeq2', 't-test.DESeq2',
                   'limma-voom')

possible.resampling <- c('loo')

possible.tests.loo = paste(possible.tests, possible.resampling, sep = '.')

# possible.resampling <- c('loo', 'bootstrap')
# 
# test.names = c()
# for(test in possible.tests){
#   for(resampling in possible.resampling){
#     test.names = c(test.names, paste(test, resampling, sep='.'))
#   }
# }
# 
# test.names.loo = test.names[grep('loo', test.names)]
# 
# test.names.important <- test.names.loo
# test.names.important <- test.names.important[-grep(c('Wilcoxon'), test.names.important)]
# test.names.important <- test.names.important[-grep(c('LRT'), test.names.important)]
# test.names.important <- test.names.important[-grep(c('t-test'), test.names.important)]
# 
# 
# 
# 
# # ------ Stability thresholds ------
# 
# top.threshols100 <- seq(100, 500, 100)
# top.threshols50 <- c(10,20,30,40,seq(50, 500, 50))
# 

# 
# 
# test.name.bo <- 'DESeq2.Wald.bootstrap'
# name.de.nofilt <- 'DESeq2.loo.nofilt'
# 

test.name <- 'DESeq2.Wald.loo'
top.n.genes <- 100

p.val.cutoffs <- c(0.1, 0.05, 0.01, 0.005, 0.01)
top.n.cutoffs = c(5, 10, 25, 50, 75, 100, 150, 200, 300, 400, 500, 1000, 2000)
top.threshols.go <- c(100, 250, 500, 1000, 2000)

example.celltypes <- names(sort(table(cao$cell.groups),decreasing=T))
example.cell.type <- example.celltypes[1]

# 
# n.resamplings <- 30
# 
# # BP, up and down
# # ordering up/down: facet_grid ggplot
# # summary: ranking 
# # supplementary tables - new plots...
# # Stability per gene: limits on axis
# 

```


# DE calculations
## LOO and bootstrap resamplings for different DE methods
```{r de-results}


if(file.exists(output.cacoa.file)){
  cao <- readRDS(output.cacoa.file) 
  cao <- Cacoa$new(cao)
}

f.changes = FALSE
for(test in possible.tests){
  # message(test)
  for(resampling in possible.resampling){
    name <- paste(test, resampling, sep='.')
    if(!(name %in% names(cao$test.results))){
      print(name)
      cao$estimateDEPerCellType(test = test, resampling.method = resampling,
                              name = name)
      f.changes = TRUE
    }
  }
}

if(f.changes) saveRDS(file = output.cacoa.file, object = cao, compress = F)


# # One run without filtration of genes with low expression
# if(!(name.de.nofilt %in% names(cao$test.results))){
#   cao$estimateDEPerCellType(test = 'DESeq2', resampling.method = 'loo',
#                                    name = name.de.nofilt, min.cell.frac = 0)
#   saveRDS(file = output.cacoa.file, object = cao, compress = F)
# }

```


## Stability: resampling with the same number of samples per case/control groups
### Calc Samples per cell types
```{r fig.width=8, fig.height=3}

n.cells.cutoff <- 50

# Number of samples is unbalanced
print(c(sum(cao$sample.groups == cao$target.level), sum(cao$sample.groups == cao$ref.level)))

# Number of cells per cell type
cnts <- cao$cell.groups %>% data.frame(anno=., group=cao$sample.per.cell[names(.)]) %>%
        table() %>% rbind() %>% t()

samples.ref <- names(cao$sample.groups)[cao$sample.groups == cao$ref.level]
samples.trgt <- names(cao$sample.groups)[cao$sample.groups == cao$target.level]
cell.types <- levels(cao$cell.groups)

samples.per.type <- c()
for(cell.type in cell.types){
  samples <- rownames(cnts)[cnts[,cell.type] > n.cells.cutoff]
  samples.per.type <- rbind(samples.per.type, 
                           c(cell.type, sum(samples %in% samples.ref), sum(samples %in% samples.trgt)))
}
samples.per.type <- as.data.frame(samples.per.type)
colnames(samples.per.type) <- c('cell.type', 'in.ref', 'in.target')
print(samples.per.type)
samples.per.type[,2:3] <- apply(samples.per.type[,2:3], 2, as.numeric)
rownames(samples.per.type) <- samples.per.type$cell.type

```


### Resampling of samples
```{r Resampling of samples, fig.width=8, fig.height=3}
n.perm.tmp <- 20
p.val.cutoff <- 0.05

df <- c()
for(n.samples.cutoff in 5:8){
  cell.types.to.work <- samples.per.type$cell.type[rowSums(samples.per.type[,2:3] < n.samples.cutoff) == 0,
                                                   drop=F]
  if(length(cell.types.to.work) == 0) next
  print(cell.types.to.work)
  samples.per.type[cell.types.to.work,]
  
  run.names <- c()
  for(i in 1:n.perm.tmp){
    run.names <- c(run.names, paste(cell.types.to.work, i, n.samples.cutoff, sep = ':'))
  }
  
  n.signif.all <- plapply(run.names, function(run.name){
    n.signif <- c()
    cell.type <- strsplit(run.name[1], ':')[[1]][1]
    samples <- rownames(cnts)[cnts[,cell.type] > n.cells.cutoff]
    samples.r <- sample(samples[samples %in% samples.ref], n.samples.cutoff)
    samples.t <- sample(samples[samples %in% samples.trgt], n.samples.cutoff)
    if(!(run.name %in% names(cao$test.results))){
      cao$estimateDEPerCellType(cell.groups = cao$cell.groups[cao$cell.groups == cell.type], 
                                name = run.name,
                                 min.cell.count = n.cells.cutoff, max.cell.count = n.cells.cutoff,
                                sample.groups = cao$sample.groups[c(samples.r, samples.t)])          
    }
    n.signif <- c(n.signif, sum(cao$test.results[[run.name]]$initial[[cell.type]]$res$padj < p.val.cutoff))
    list(n.signif = n.signif, samples.r=samples.r, samples.t=samples.t)
  }, n.cores=min(length(run.names), 30) )
  names(n.signif.all) <- run.names
  
  for(run.name in run.names){
    cell.type <- strsplit(run.name[1], ':')[[1]][1]
    df <- rbind(df, cbind(n.signif.all[[run.name]]$n.signif, cell.type, n.samples.cutoff))
  }
}
df <- as.data.frame(df)
colnames(df) <- c('n.signif', 'cell.type', 'n.samples.cutoff')
df$n.signif <- as.numeric(df$n.signif)

saveRDS(file = output.cacoa.file, object = cao, compress = F)
saveRDS(file = output.df.file, object = df, compress = F)



```


### Plotting
```{r, fig.width=8, fig.height=3}


p.n.signif <- ggplot(df, aes(x = cell.type, y = n.signif, color = n.samples.cutoff)) + 
geom_boxplot(outlier.size=0, notch=FALSE) + geom_point(position = position_jitterdodge()) + theme_minimal() + scale_y_log10()

p.n.signif

pdf(paste(path.figures,  dataset.name,
          '_number_of_significant_genes.pdf', sep = ''), 
    width = 7, height = 3)
print(p.n.signif)
dev.off()

```


## Fixation of # of cells per sample
```{r}
names.fixed.cells = c('deFixedNumCell100', 'deFixedNumCell50')

if(!(names.fixed.cells[1] %in% names(cao$test.results))){
  cao$estimateDEPerCellType(max.cell.count=100, min.cell.count = 100, 
                            name='deFixedNumCell100', resampling.method='loo')  
  saveRDS(file = output.cacoa.file, object = cao, compress = F)
}

if(!(names.fixed.cells[2] %in% names(cao$test.results))){
  cao$estimateDEPerCellType(max.cell.count=50, min.cell.count = 50, 
                            name='deFixedNumCell50', resampling.method='loo')
  saveRDS(file = output.cacoa.file, object = cao, compress = F)
}


```


## LOO within LOO for DESeq2.Wald
```{r loo-within-loo}
test <- 'DESeq2.Wald'
sample.names <- names(cao$sample.groups)
f.changes = FALSE
for(s in sample.names){
      sample.groups.tmp <- cao$sample.groups
      sample.groups.tmp <- sample.groups.tmp[!(names(sample.groups.tmp) %in% s)]
      name.tmp <- paste('remove', s, sep='.')
      if(name.tmp %in% names(cao$test.results)) next
      f.changes = TRUE
      cao$estimateDEPerCellType(test = test, resampling.method = 'loo',
                                name = name.tmp,
                                sample.groups = sample.groups.tmp)
}

if(f.changes) saveRDS(file = output.cacoa.file, object = cao, compress = F)


# summarize new "p-values"


summ.resamplings <- list(initial=cao$test.results$DESeq2.Wald.loo$initial)
cell.types <- names(summ.resamplings$initial)


for(s in sample.names){
  name.tmp <- paste('remove', s, sep='.')
  print(name.tmp)
  res.tmp <-cao$test.results[[name.tmp]]$initial
  for(cell.type in names(res.tmp)){
    stab.sum.rank = rep(0, nrow(res.tmp[[cell.type]]$res))
    if(length(stab.sum.rank) == 0) next
    g = rownames(res.tmp[[cell.type]]$res)
    for(smpl in setdiff(names(cao$test.results[[name.tmp]]), 'initial')){
      if(is.null(cao$test.results[[name.tmp]][[smpl]][[cell.type]])) next
      stab.sum.rank = stab.sum.rank + 
        rank(cao$test.results[[name.tmp]][[smpl]][[cell.type]][g, 'pvalue'])
    }
    res.tmp[[cell.type]]$res$pvalue <- stab.sum.rank
    res.tmp[[cell.type]] = res.tmp[[cell.type]]$res
  }
  summ.resamplings[[s]] <- res.tmp
} 
cao$test.results[['loo.within.loo']] <- summ.resamplings
```


## Covariates (MS dataset only)
```{r}
# 
# if((dataset.name == 'MS') & !('DESeq2.Wald.PMI' %in% names(cao$test.results) )){
#   ms.cov <- read.table('/d0/home/aigolkina/coda/cacoa-examples/report/ms_covariate.csv', 
#                       sep = ',', header = 1)
#   rownames(ms.cov) <- ms.cov[,1]
#   
#   
#   sample_metadata <- "/d0/home/aigolkina/coda/cacoa-examples/report/ms_metadata.csv" %>%
#     read.table(sep = ',', header = 1) %>% mutate(sample=substr(.$cell, 20, 100)) %>% split(.$sample) %>%
#     lapply(`[`, 1,) %>% do.call(rbind, .) %>% .[,5:14]
#   # sample_metadata %<>% lapply(setNames, .$sample)
#   sample_metadata <- as.data.frame(sample_metadata)
#   rownames(sample_metadata) <- sample_metadata$sample
#   
#   ms.cov$batch = sample_metadata[rownames(ms.cov), 'Seqbatch'] %in% c('SEQ1', 'SEQ3')
#   ms.cov$disease <- sample_metadata[rownames(ms.cov), 'diagnosis'] %in% c('MS')
# 
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald',
#                           covariates = ms.cov[names(cao$sample.groups),'batch', drop=F],
#                           name='DESeq2.Wald.batch', resampling.method = 'loo')
#   
#   # cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald',
#   #                         name='xxx', resampling.method = 'loo', n.cores=1)
# 
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald',
#                           covariates = ms.cov[names(cao$sample.groups),'PMI', drop=F],
#                           name='DESeq2.Wald.PMI', resampling.method = 'loo')
#   
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', max.resamplings = 1,
#                           covariates = ms.cov[names(cao$sample.groups),'Age', drop=F],
#                           name='DESeq2.Wald.Age', resampling.method = 'loo')
#   
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', max.resamplings = 1,
#                           covariates = ms.cov[names(cao$sample.groups),'Sex', drop=F],
#                           name='DESeq2.Wald.Sex', resampling.method = 'loo')
#   
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', max.resamplings = 1,
#                           covariates = ms.cov[names(cao$sample.groups),'RIN', drop=F],
#                           name='DESeq2.Wald.RIN', resampling.method = 'loo')
#   
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', max.resamplings = 1,
#                           covariates = ms.cov[names(cao$sample.groups),c('RIN', 'PMI', 'Age', 'Sex'), drop=F],
#                           name='DESeq2.Wald.all.cov', resampling.method = 'loo')
# 
#   cao$estimateDEPerCellType(independent.filtering=TRUE, test='DESeq2.Wald', max.resamplings = 1,
#                             name = 'de')
#   
#   saveRDS(file = output.cacoa.de.file.de, object = cao, compress = F)
# }


```



# DE Stability plots
## p-values: shuffling groups. Each line - cell type
```{r, fig.width = 2.5, fig.height = 3}
shuffle.pref <- 'de.shuffling'
# n.shullfe <- 30
n.shullfe <- 100
if(!(paste0(shuffle.pref, 'edgeR', 1, collapse = '')) %in% names(cao$test.results)){
  for(i in 1:n.shullfe) {
    print(i)
    set.seed(i+239)
    sample.groups.shuffling <- sample(cao$sample.groups)
    names(sample.groups.shuffling) <- names(cao$sample.groups)
    print(sample.groups.shuffling)
    for(test in c('DESeq2', 'edgeR', 'Wilcoxon.DESeq2')){
    # for(test in c('edgeR')){
      name <- paste0(shuffle.pref, test, i, collapse = '')
      cao$estimateDEPerCellType(name = name, 
                                sample.groups = sample.groups.shuffling, test = test,
                                n.cores=80)  
    }
  }
  saveRDS(file = output.cacoa.file, object = cao, compress = F)
}

p.val.steps = exp(seq(-6, 0, by = 0.1) * log(10))
df.summary.tests <- c()
for(test in c('DESeq2', 'Wilcoxon.DESeq2',  'edgeR')){
  print(test)
# for(test in c( 'edgeR')){
  df <- c()
  for(i in 1:n.shullfe) {
    print(i)
    name <- paste0(shuffle.pref, test, i, collapse = '')
    for(cell.type in names(cao$test.results[[name]]$initial)){
      pval.cell.type = cao$test.results[[name]]$initial[[cell.type]]$res$pvalue
      n.res.cell.type = length(pval.cell.type)
      
      tmp <- sapply(p.val.steps, function(p) { return(sum(pval.cell.type < p))})
      
      df.tmp <- data.frame(cell.type = cell.type, 
                           pval = p.val.steps, 
                           frac = tmp / n.res.cell.type,
                           exp.num = p.val.steps * n.res.cell.type,
                           num=tmp)
      df <- rbind(df, df.tmp)

    }
  }
  df = df[df$num > 1, ]
  # get mean values across sufflings
  df.summary <- df %>% group_by(interaction(cell.type, pval)) %>%
    summarize(frac.mean = mean(frac), frac.med = median(frac), num.med = median(num), 
              cell.type = cell.type[1], pval = pval[1])
  df.summary$ratio.mean = df.summary$frac.mean / df.summary$pval
  df.summary$ratio.med = df.summary$frac.med / df.summary$pval
  df.summary[df.summary == 0] <- NA
  df.summary$num.ceil <- as.factor(ceiling(log(df.summary$num.med, base=10)))
  df.summary$num.floor <- as.factor(floor(log(df.summary$num.med, base=10)))
  df.summary$test <- test
  
  
  df.summary <- df.summary[!is.na(df.summary$ratio.med),]
  
  df.summary.tests <- rbind(df.summary.tests, df.summary)
}

# p.median.horiz <- ggplot(df.summary.tests, aes(x = pval, y = ratio.med, colour = num.med, group = cell.type))

df.summary.tests$test[df.summary.tests$test == 'Wilcoxon.DESeq2'] = 'Wilcoxon'

p.median.horiz <- ggplot(df.summary.tests[df.summary.tests$num.ceil != 0,], 
                         aes(x = pval, y = ratio.mean, colour = num.ceil, group = cell.type))+ 
  geom_line(alpha=0.5) + 
  # geom_point(alpha=0.5) +
  cao$plot.theme +
  # theme(legend.title = element_blank()) +
  
  scale_x_continuous(trans='log10', limits = c(10^-4, 1),
                     breaks = c(10^-4, 10^-3, 10^-2, 10^-1, 1), 
                     labels=c('1e-4', '1e-3', '1e-2', '1e-1', 1)) +
  scale_y_continuous(trans='log10', breaks = c(10^-1, 0, 10^1)) + geom_hline(yintercept=1, linetype="dashed", 
                color = "black") + xlab('p-value cutoff') + ylab('Mean of observed/expected') +
  guides(color=guide_legend(ncol=4, title.position = "top", title.vjust = 1))  + 
  theme(legend.position="bottom", legend.title=element_text(size=7), 
    legend.text=element_text(size=6)) +
  scale_color_discrete(name = "Round log10 # DE genes") +
  # scale_color_manual(labels = c(">10", ">100", ">1e3", '>1e4', '>1e5'), values = c(1, 2, 3, 4, 5))+
  scale_color_viridis(discrete = TRUE, name = "Num. of DE genes",
                      labels = c(">10", ">100", ">1000", '>10000'), breaks = c(1, 2, 3, 4)) +
  facet_grid(rows = vars(test)) + theme(strip.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA)) +
  coord_cartesian(ylim = c(10^-2,10^2), expand=0)



p.median.horiz 

pdf(paste(path.figures,  dataset.name,
          '_p_values_shuffling_groups_median_all2.pdf', sep = ''),
    width = 2.2, height = 3)
print(p.median.horiz)
dev.off()
```

```{r, fig.width = 6, fig.height = 3}
p.median.horiz <- ggplot(df.summary.tests[df.summary.tests$num.ceil != 0,], 
                         aes(x = pval, y = ratio.mean, colour = num.ceil, group = cell.type))+ 
  geom_line(alpha=0.5) + 
  # geom_point(alpha=0.5) +
  cao$plot.theme +
  # theme(legend.title = element_blank()) +
  
  scale_x_continuous(trans='log10', limits = c(10^-4, 1),
                     breaks = c(10^-4, 10^-3, 10^-2, 10^-1, 1), 
                     labels=c('1e-4', '1e-3', '1e-2', '1e-1', 1)) +
  scale_y_continuous(trans='log10', breaks = c(10^-1, 0, 10^1)) + geom_hline(yintercept=1, linetype="dashed", 
                color = "black") + xlab('p-value cutoff') + ylab('Mean of observed/expected') +
  guides(color=guide_legend(ncol=4, title.position = "top", title.vjust = 1))  + 
  theme(legend.position="bottom", legend.title=element_text(size=7), 
    legend.text=element_text(size=6)) +
  scale_color_discrete(name = "Round log10 # DE genes") +
  # scale_color_manual(labels = c(">10", ">100", ">1e3", '>1e4', '>1e5'), values = c(1, 2, 3, 4, 5))+
  scale_color_viridis(discrete = TRUE, name = "Num. of DE genes",
                      labels = c(">10", ">100", ">1000", '>10000'), breaks = c(1, 2, 3, 4)) +
  facet_grid(cols = vars(test)) + theme(strip.background = element_blank(),
        panel.border = element_rect(colour = "black", fill = NA)) +
  coord_cartesian(ylim = c(10^-2,10^2), expand=0)


p.median.horiz 


pdf(paste(path.figures,  dataset.name,
          '_p_values_shuffling_groups_median_all2_horiz.pdf', sep = ''),
    width = 6, height = 3)
print(p.median.horiz)
dev.off()
```



```{r, fig.width = 5, fig.height = 5}






# 
# saveRDS(df.summary, paste(path.figures,  dataset.name,
#           '_p_values_shuffling_groups_median_', test, '.rds', sep = ''))

```


## Stability of cell types (one DE method is used)
```{r, fig.width=4, fig.height=3}
cao$estimateDEStabilityPerCellType(top.n.genes = top.n.genes,
                                   de.name = test.name,
                                   name = 'de.stab.filt', )

# cao$estimateDEStabilityPerCellType(top.n.genes = top.n.genes,
#                                    de.name = name.de.nofilt, 
#                                    name = 'de.stab.nofilt')
# 
# cao$plotDEStabilityPerCellType(sort.order = T, name = 'de.stab.nofilt')  + 
#   ylim(0.1, 0.8)

p <- cao$plotDEStabilityPerCellType(name = 'de.stab.filt', notch = T, set.fill = F) 
p 

```


```{r, fig.width=4, fig.height=3}
pdf(paste(path.figures,  dataset.name,
          '_stab_per_cell_types.pdf', sep = ''), 
    width = 4, height = 2.7) 
print(p)
dev.off()

```



## Trend: One DE method and Different thresholds
### N top genes
```{r, fig.width=4, fig.height=3}


cao$estimateDEStabilityTrend(de.name = test.name, 
                             top.n.genes = c(top.n.cutoffs),
                             name = 'de.stab.trend.topn')
p <- cao$plotDEStabilityTrend(name = 'de.stab.trend.topn', jitter.alpha = 0.1) + xlab('Top N genes') +
   scale_fill_viridis(discrete = TRUE) 
p 

pdf(paste(path.figures,  dataset.name, '_trend_top_n.pdf', sep = ''), 
    width = 4, height = 3) 
print(p)
dev.off()

```

###p-values
```{r, fig.width=1.8, fig.height=3}
cao$estimateDEStabilityTrend(de.name = test.name, p.val.cutoffs = p.val.cutoffs,
                             name = 'de.stab.trend.pval') 
p <- cao$plotDEStabilityTrend(name = 'de.stab.trend.pval', jitter.alpha = 0.5) + xlab('p-value cutoffs') +
   scale_fill_viridis(discrete = TRUE) 
p

pdf(paste(path.figures,  dataset.name, '_trend_pval.pdf', sep = ''),
    width = 1.8, height = 3)
print(p)
dev.off()

```


## Stability per gene
### Smooth "per gene" stability
```{r, fig.width=5.2, fig.height=3}

plot.list <- list()
for(t.n.genes in c(100, 200, 500, 1000)){
  plot.list[[length(plot.list) + 1]] <- cao$estimateDEStabilityPerGene(de.name = test.name, 
                                                         visualize=T, top.n.genes = t.n.genes) +
    ggtitle(paste0(c('Top', t.n.genes, 'genes'), collapse = ' ')) 
}

cao$est

# ggarrange(plots=plot.list, nrow=2)
p <- plot.list[[1]]
p

# pdf(paste(path.figures,  dataset.name, '_per_gene_smooth.pdf', sep = ''), 
#     width = 5.2, height = 3) 
# print(p)
# dev.off()
```
### Rank-based stability
```{r}

# Variance of rank: LOO
cell.type = levels(cao$cell.groups)[1]
cao$plotDEStabilityPerGene(name = test.name,
                       cell.type = example.cell.type,
                       stability.score = 'stab.var.rank')

# # Median rank: bootstrap
# cao$plotDEStabilityPerGene(name = test.name.bo,
#                        cell.type = example.cell.type,
#                        stability.score = 'stab.median.rank')

# Median rank: LOO
cao$plotDEStabilityPerGene(name = test.name,
                       cell.type = example.cell.type,
                       stability.score = 'stab.median.rank')
```



### Stability of genes on Volcano
```{r, fig.height=5, fig.width=5}

# cao$estimateDEStabilityPerGene(de.name = test.name, visualize = F)
# cao$plotVolcano(name = test.name, color.var = 'Stability')


```


# Update cacoa object 
```{r}
suppressMessages(devtools::load_all("/home/aigolkina/coda/cacoa/"))
cao <- Cacoa$new(cao)
```



## Stability for different tests
Methods are sorted by means
```{r, fig.width=8, fig.height=4}
tests.for.stability = c("DESeq2.Wald.loo", "DESeq2.LRT.loo", "edgeR.loo", "Wilcoxon.DESeq2.loo",
                        "t-test.DESeq2.loo", "limma-voom.loo", "loo.within.loo" ,
                        'ga.IN-PV_res8_ms_cells_')
                        # 'ga.Macrophages:1')
names.for.stability = c("DESeq2-Wald", "DESeq2-LRT", "edgeR", "Wilcoxon test",
                        "t-test", "limma-voom", "LOO: DESeq2", 'Genetic Algorithm')
names(names.for.stability) <- tests.for.stability
colors.for.stability <- c('#C2B8A3', '#A19882', '#8FC1D4', '#C9CCD5', '#B97A95', '#FFE3E3', '#B24080')
names(colors.for.stability) <- c("t-test.DESeq2.loo", "Wilcoxon.DESeq2.loo", "limma-voom.loo",  "edgeR.loo",
                                 "DESeq2.Wald.loo", "DESeq2.LRT.loo",  "loo.within.loo")


plot.list <- list()
jacc.all <- c()
# for(t.n.genes in c(100, 250, 500)){
for(t.n.genes in c(100)){
  # cao$estimateDEStabilityPerTest(de.names = c(test.names.loo, name.de.nofilt, 
  #                                             'DESeq2.Wald.PMI', 'DESeq2.Wald.Age',
  #                                             'DESeq2.Wald.Sex', 'DESeq2.Wald.RIN',
  #                                             'DESeq2.Wald.all.cov', 'deFixedMaxNumCell',
  #                                             'deFixedNumCell100'),
  #                                top.n.genes = t.n.genes)
  
  cao$estimateDEStabilityPerTest(de.names = tests.for.stability,
                                 top.n.genes = t.n.genes)

  jacc.all <- rbind(jacc.all, cbind(cao$test.results$jacc.per.test, t.n.genes))
  
  p1 <- cao$plotDEStabilityPerTest(sort.order = T) + ggtitle(paste('Top', t.n.genes, 'genes'))
  plot.list[[length(plot.list) + 1]] <- p1 + scale_x_discrete(labels = names.for.stability) +
     scale_fill_manual(values = colors.for.stability)
}

p.val.cutoff = 0.05
cao$estimateDEStabilityPerTest(de.names = setdiff(tests.for.stability, c('Wilcoxon.DESeq2.loo', 'ga.IN-PV_res8_ms_cells_', 'loo.within.loo')),
                               p.val.cutoff = p.val.cutoff, name = 'jacc.per.test.pval')

plot.list[[length(plot.list) + 1]] <- 
  cao$plotDEStabilityPerTest(sort.order = T, name = 'jacc.per.test.pval') + 
  ggtitle(paste('P.adj cutoff', p.val.cutoff)) + scale_x_discrete(labels = names.for.stability) +
  scale_fill_manual(values = colors.for.stability)

p <- ggarrange(plots=plot.list[c(1, length(plot.list))], ncol=2)

p
# pdf(paste(path.figures,  dataset.name, '_tests_cutoffs.pdf', sep = ''),
#     width = 8, height = 4)
# print(p)
# dev.off()


```





### Saving
```{r, fig.width=8, fig.height=4}
p1.ntop = plot.list[[1]]
p1.pval = plot.list[[length(plot.list)]]


# saveRDS(p1.ntop, paste(path.figures,  dataset.name, '_tests_cutoffs_100top.rds', sep = ''), compress = F)
# saveRDS(p1.pval, paste(path.figures,  dataset.name, '_tests_cutoffs_pval.rds', sep = ''), compress = F)

# plot.list[[1]] 


pdf(paste(path.figures,  dataset.name, '_tests_cutoffs_100top.pdf', sep = ''),
    width = 5, height = 4)
print(p1.ntop)
dev.off()


pdf(paste(path.figures,  dataset.name, '_tests_cutoffs_pval.pdf', sep = ''),
    width = 2.5, height = 4)
print(p1.pval)
dev.off()

```

Means
```{r, fig.width=5, fig.height=3.5}
jacc.all.m <- jacc.all %>% group_by(interaction(group, t.n.genes) ) %>%
  summarize(m = mean(value), group = group[1], t.n.genes = t.n.genes[1])

pl <- ggplot(jacc.all.m, aes(group, m, color = as.factor(t.n.genes), 
                                group = as.factor(t.n.genes) )) + geom_line()  + theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(color = 'top N genes') +
  ylab('Mean Jaccard index') + xlab('')
pl
```

Medians
```{r, fig.width=5, fig.height=3.5}
jacc.all.m <- jacc.all %>% group_by(interaction(group, t.n.genes)) %>%
  summarize(median = median(value), group = group[1], t.n.genes = t.n.genes[1])

pl <- ggplot(jacc.all.m, aes(group, median, color = as.factor(t.n.genes), 
                                group = as.factor(t.n.genes) )) + geom_line()  + theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) + labs(color = 'top N genes') +
  ylab('Median Jaccard index') + xlab('')
pl
```


## Stability across datasets
```{r, fig.width=5, fig.height=3.5}
test.jac <- sapply(unique(cao$test.results$jacc.per.test$group), 
       function(s){
         mean(cao$test.results$jacc.per.test$value[cao$test.results$jacc.per.test$group == s])
       })
test.jac

test.jac.log <- read.table(methods.cmp.file)
test.jac.log <- test.jac.log[test.jac.log$V1 != dataset.name,]
test.jac.log = rbind(test.jac.log, 
                     data.frame(V1=dataset.name,
                                V2=names(test.jac),
                                V3=test.jac) )


write.table(x = test.jac.log, 
            file = methods.cmp.file,
            quote = F, col.names = F, row.names = F)

```


```{r, fig.width=5, fig.height=4}
test.jac.log <- read.table(methods.cmp.file)
# test.jac.log <- unique(test.jac.log)

tests.for.stability = c("DESeq2.Wald.loo", "DESeq2.LRT.loo", "edgeR.loo", "Wilcoxon.DESeq2.loo",
                        "t-test.DESeq2.loo", "limma-voom.loo", "loo.within.loo")
names.for.stability = c("DESeq2-Wald", "DESeq2-LRT", "edgeR", "Wilcoxon test",
                        "t-test", "limma-voom", "LOO: DESeq2")
names(names.for.stability) <- tests.for.stability


test.jac.log <- test.jac.log[test.jac.log$V2 %in% tests.for.stability, ]

datasets.cols <- c('#e45414', '#efa66c', '#7a2413', '#38A3A5', '#1E5128', '#4E9F3D')


test.jac.log = test.jac.log[test.jac.log$V1 != 'hc',]
test.jac.log = test.jac.log[rowSums(is.na(test.jac.log)) == 0,]

p <- ggplot(test.jac.log, aes(reorder(V2, V3), V3, group = V2 )) + 
  # geom_boxplot(notch = F, outlier.shape = NA) + theme_minimal() +
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  ylab('Mean Jaccard index') + xlab('') + theme(legend.position = "right") + 
  theme(plot.title = element_text(size = 10)) + 
  # geom_jitter(aes(color = V1), width = 0.2) + 
  labs(color='Datasets') +
  geom_line(aes(color = V1, group=V1), width = 0.2) +
  geom_point(aes(x = V2, y=V3, color = V1)) +
  scale_color_manual(values = datasets.cols,
                     labels = c("Epilepsy", "Autism", "Alzheimer", 'MS', 'SCC', 'PF'), 
                     breaks = c("ep", "autism", "alzheimer1", 'ms', 'scc', 'pf')) + 
  scale_x_discrete(labels = names.for.stability)


  
p
```


```{r, fig.width=5, fig.height=3.5}
pdf(paste(path.figures,  dataset.name, '_tests_datasets.pdf', sep = ''), 
    width = 4, height = 3) 
print(p)
dev.off()

```



## Stability between tests
```{r, fig.width=6, fig.height=6}
# cao$estimateDEStabilityBetweenTests(de.names = c(test.names.loo, name.de.nofilt),
#                                     top.n.genes = top.n.genes)

cao$estimateDEStabilityBetweenTests(de.names = c(possible.tests.loo, 'ga.OPC_res8_ms_cells_', 'tmp.tmp'),
                                    top.n.genes = top.n.genes)

# cao$estimateDEStabilityBetweenTests(de.names = test.names.loo,
#                                     top.n.genes = top.n.genes)

p <- cao$plotDEStabilityBetweenTests() 
p

pdf(paste(path.figures,  dataset.name, '_tests_between.pdf', sep = ''), 
    width = 6, height = 6) 
print(p)
dev.off()

```

# Update cacoa object 
```{r}
suppressMessages(devtools::load_all("/home/aigolkina/coda/cacoa/"))
cao <- Cacoa$new(cao)
```



# GO calculations
## ORA: over-representation analysis

```{r gsea, message=FALSE, warning=FALSE}

test.enrichment = 'DESeq2.Wald.loo'
calc.mode = 'go'

test.enrichment.tmp = paste(test.enrichment, calc.mode, sep='.')
cao$test.results[[test.enrichment.tmp]] = cao$test.results[[test.enrichment]]
test.enrichment = test.enrichment.tmp

if(calc.mode %in% c('go')) {  # if

  org <- org.Hs.eg.db
  cao$getGOEnvironment(org.db = org)

  # for(top.tresh in top.threshols.go) {
  for(top.tresh in c(250,500,1000,2000)) {
    print(top.tresh)    
    go.pref = paste('go', test.enrichment, top.tresh, sep = '.')

  
    pref = paste(go.pref, 'init', sep = '.')
    cao$estimateOntology(type = 'GO', org.db=org, de.name = test.enrichment, name = pref,
                         n.top.genes = top.tresh)
  
    # Save initial values of DE analysis
    cao$test.results[[test.enrichment]]$save.init <- cao$test.results[[test.enrichment]]$initial
  
    # Names of resamplings
    prefs.resampling <- setdiff(names(cao$test.results[[test.enrichment]]), c('initial', 'save.init'))
    prefs.done = names(cao$test.results)
    print(prefs.resampling)
  
    # This loop is required due to the "random" Error:
    # for(irep in 1:4){
    

    # res <- plapply(prefs.resampling, function(pref.subs) {  # Parallel by resamplings
    res = list()
    for(pref.subs in prefs.resampling){
      print(pref.subs)
      
      # for(pref.subs in prefs.resampling) {
      pref.subs.res <- paste(calc.mode, pref.subs, sep = '.')
    
    
      # Replace values for each resampling
      cao$test.results[[test.enrichment]]$initial = list()
      for(cell.type in names(cao$test.results[[test.enrichment]][[pref.subs]])){
        cao$test.results[[test.enrichment]]$initial[[cell.type]]$res <-
          cao$test.results[[test.enrichment]][[pref.subs]][[cell.type]]
        # Add column Gene
        cao$test.results[[test.enrichment]]$initial[[cell.type]]$res$Gene <-
          rownames(cao$test.results[[test.enrichment]]$initial[[cell.type]]$res)
      }
      res[[pref.subs]] <- cao$estimateOntology(type = 'GO', org.db=org, de.name = test.enrichment,
                           name = pref.subs.res, n.top.genes = top.tresh)

      # return(cao$test.results[[pref.subs.res]])
    }
    # }, n.cores=30, progress=TRUE)

    # names(res) <- prefs.resampling
    
    pref.resampling = paste(go.pref, 'res', sep = '.')
    res[['initial']] = cao$test.results$go.DESeq2.Wald.loo.go.100.init
    cao$test.results[[pref.resampling]] = res

  
    # Return initial values back
    
    cao$test.results[[test.enrichment]]$initial <- cao$test.results[[test.enrichment]]$save.init
  
    saveRDS(object = cao, file = output.cacoa.file, compress = F)
    
  }
}

```

## GSEA calculation
```{r}

test.enrichment = 'DESeq2.Wald.loo'
calc.mode = 'gsea'

test.enrichment.tmp = paste(test.enrichment, calc.mode, sep='.')
cao$test.results[[test.enrichment.tmp]] = cao$test.results[[test.enrichment]]
test.enrichment = test.enrichment.tmp

if(calc.mode %in% c('gsea')) {  # if

  org <- org.Hs.eg.db
  cao$getGOEnvironment(org.db = org)

  go.pref = paste('gsea', test.enrichment, sep = '.')

  pref = paste(go.pref, 'init', sep = '.')
  cao$estimateOntology(type = 'GSEA', org.db=org, de.name = test.enrichment, name = pref)

  # Save initial values of DE analysis
  cao$test.results[[test.enrichment]]$save.init <- cao$test.results[[test.enrichment]]$initial

  # Names of resamplings
  prefs.resampling <- setdiff(names(cao$test.results[[test.enrichment]]), c('initial', 'save.init'))
  prefs.done = names(cao$test.results)
  print(prefs.resampling)

  # This loop is required due to the "random" Error:
  # for(irep in 1:4){
  
  # res <- plapply(prefs.resampling, function(pref.subs) {  # Parallel by resamplings
  res = list()
  for(pref.subs in prefs.resampling){
    print(pref.subs)
    
    # for(pref.subs in prefs.resampling) {
    pref.subs.res <- paste(calc.mode, pref.subs, sep = '.')
  
  
    # Replace values for each resampling
    cao$test.results[[test.enrichment]]$initial = list()
    for(cell.type in names(cao$test.results[[test.enrichment]][[pref.subs]])){
      cao$test.results[[test.enrichment]]$initial[[cell.type]]$res <-
        cao$test.results[[test.enrichment]][[pref.subs]][[cell.type]]
      # Add column Gene
      cao$test.results[[test.enrichment]]$initial[[cell.type]]$res$Gene <-
        rownames(cao$test.results[[test.enrichment]]$initial[[cell.type]]$res)
    }
    res[[pref.subs]] <- cao$estimateOntology(type = 'GSEA', org.db=org, de.name = test.enrichment,
                         name = pref.subs.res)

    # return(cao$test.results[[pref.subs.res]])
  }
  # }, n.cores=30, progress=TRUE)

  # names(res) <- prefs.resampling
  
  pref.resampling = paste(go.pref, 'res', sep = '.')
  res[['initial']] = cao$test.results[[pref]]
  cao$test.results[[pref.resampling]] = res


  # Return initial values back
  
  cao$test.results[[test.enrichment]]$initial <- cao$test.results[[test.enrichment]]$save.init

  saveRDS(object = cao, file = output.cacoa.file, compress = F)

}


```

# Update cacoa object 
```{r}
suppressMessages(devtools::load_all("/home/aigolkina/coda/cacoa/"))
cao <- Cacoa$new(cao)
```




# Go Stability plots
## ORA
```{r, fig.height=13, fig.width=6}

test.name = "DESeq2.Wald.loo"
plot.list <- list()
go.median <- c()
for(t.n.genes in top.threshols.go){
  go.result = paste('go', test.name, 'go', t.n.genes , 'res', sep = '.')
  cao$estimateGOStabilityPerCellTypePerm(go.result = go.result, padj.go = 0.05)
  
  tmp <- cao$test.results$go.jaccards %>% group_by(group) %>% 
    summarize(median = mean(value, ), group = group[1])
  
  go.median <- rbind(go.median, cbind(tmp, t.n.genes))

}



```

# GSEA
```{r}

go.pref = paste('gsea', test.enrichment, sep = '.')

gsea.result = paste('gsea', test.name, 'gsea' , 'res', sep = '.')
cao$estimateGSEAStabilityPerCellTypePerm2(go.result = gsea.result, padj.go = 0.05)

tmp <- cao$test.results$go.jaccards %>% group_by(group) %>% 
  summarize(median = mean(value, ), group = group[1])

t.n.genes = 'gsea'
gsea.median <- cbind(tmp, t.n.genes)

```





## Plot GO stability
```{r, fig.height=3, fig.width=4}
x_order <- c('100', '250', '500', '1000', '2000', 'gsea')

names.go.stab = c("GO:100", 'GO:250','GO:500','GO:1000','GO:2000', 'GSEA')
names(names.go.stab) <- x_order

go.colors <- c('#1E5128', '#4E9F3D', '#6ECB63', '#B1E693', '#D8E9A8', '#ED8E7C')
names(go.colors) <- x_order

p <- ggplot(rbind(go.median, gsea.median), 
            aes(factor(as.factor(t.n.genes), level = x_order), median, 
                                group = as.factor(t.n.genes),  fill=as.factor(t.n.genes) )) + 
  geom_boxplot(notch=F, outlier.shape=NA)  + theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))  + 
  ylab('Jaccard Index') + geom_jitter(width = 0.2, alpha = 0.5) + xlab('') +
  theme(legend.position = "none") +
  scale_fill_manual(values = go.colors) + scale_x_discrete(labels = names.go.stab)  + ggtitle('Ontology stability') + ylim(0, 1)
p 

pdf(paste(path.figures,  dataset.name, '_go_stab.pdf', sep = ''),
    width = 3.5, height = 4)
print(p)
dev.off()
```


