---
title: 'Preprocess all datasets with Pagoda2 and Conos'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

## Libraries

```{r libs, message=FALSE}
library(tidyverse)
library(magrittr)
library(sccore)
library(pagoda2)
library(conos)
library(dataorganizer)
library(Matrix)
library(reticulate)

devtools::load_all()

N_CORES <- 45
FORCE <- FALSE
```

```{r, message=FALSE}
virtualenv_create("r-scrublet")
virtualenv_install("r-scrublet", c("scrublet", "matplotlib"))
use_virtualenv("r-scrublet")
mpl <- import("matplotlib")
mpl$use('Agg') # Otherwise it shows Qt error in RStudio
scrublet <- import("scrublet")

estimateDoubletInfo <- function(mats, progress=FALSE) {
  dub.info <- sccore::plapply(mats, function(m) {
    suppressMessages(scrublet$Scrublet(t(m), random_state=as.integer(42))$scrub_doublets()) %>%
      lapply(setNames, colnames(m))
  }, n.cores=1, progress=progress)

  lapply(c(scores=1, mask=2), function(i) {
    lapply(dub.info, `[[`, i) %>% unname() %>% unlist()
  })
}
```

## Alzheimer

Cells are already filtered by mit. fraction, and doublets are removed. We filter only by minimum 500 UMIs per cell and scrublet scores.

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('ASD/con.rds'), function() {
  mat <- DataPath("AZ/cell_counts.csv") %>% data.table::fread(sep=",") %>%
    {set_rownames(mltools::sparsify(.[, 2:ncol(.)]), .$V1)} %>%
    .[rowSums(. > 0) >= 10,]

  cell.metadata <- DataPath("AZ/cell_metadata.csv") %>% read_delim(delim='\t') %>%
    rename(cell=sampleID, sample=patient) %>%
    select(cell, batch, sample, sex, cellType, subclustID) %>%
    filter(!grepl("un", sample), !(cellType %in% c('doublet', 'unID')))

  sample.metadata <- group_by(cell.metadata, batch, sample, sex) %>%
    summarise(n=n()) %>% select(-n) %>% lapply(setNames, .$sample)

  cell.metadata %<>% lapply(setNames, .$cell)

  mat.per.samp <- splitMatrixByFactor(mat, cell.metadata$sample)

  dub.info <- estimateDoubletInfo(mat.per.samp)
  p2s <- plapply(mat.per.samp, createPagoda, min.transcripts.per.cell=500, dub.scores=dub.info$scores,
                 dub.threshold=0.3, mc.preschedule=TRUE, n.cores=N_CORES, progress=FALSE)

  createConos(p2s, sample.meta=sample.metadata, cell.meta=cell.metadata, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.2, alpha=0.2)
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```

## Autism

Cells are already filtered by mitochondrial fraction of 0.05 and UMI threshold ~500, no additional filtration is needed.

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('ASD/con.rds'), function() {
  mat <- Seurat::Read10X(DataPath("ASD")) %>% .[rowSums(. > 0) >= 10,]
  meta <- read_delim(DataPath("ASD/meta.txt"), delim='\t') %>%
    rename(cellType=cluster, PMI=`post-mortem interval (hours)`) %>%
    mutate(cellType=gsub("-I(I)?", "", cellType))

  sample.metadata <- meta %>%
    group_by(sample, individual, region, age, sex, diagnosis, Capbatch, Seqbatch) %>%
    summarise(PMI=median(PMI)) %>% lapply(setNames, .$sample)

  sample.metadata$region_hr <- sample.metadata$sample %>% strsplit('_') %>% sapply(`[[`, 2)

  cell.metadata <- meta %>% lapply(setNames, .$cell)

  mat.per.samp <- splitMatrixByFactor(mat, cell.metadata$sample)
  mat.per.cap <- splitMatrixByFactor(mat, cell.metadata$Capbatch)

  dub.info <- estimateDoubletInfo(mat.per.cap)
  p2s <- plapply(mat.per.samp, createPagoda, dub.scores=dub.info$scores, dub.threshold=0.17,
                 mc.preschedule=TRUE, n.cores=N_CORES, progress=FALSE)
  createConos(p2s, sample.meta=sample.metadata, cell.meta=cell.metadata, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.1, alpha=0.1)
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```

## Epilepsy

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('EP/con.rds'), function() {
  con.pap <- read_rds(DataPath("EP/con_filt_samples.rds")) %>% conos::Conos$new()
  cell.metadata <- DataPath("EP/annotation.csv") %>% read_csv() %>%
    rename(cellType=l4) %>% lapply(setNames, .$cell)
  sample.metadata <- DataPath("EP/sample_info.csv") %>% read_csv() %>% lapply(setNames, .$Alias)

  con.pap$samples %>% lapply(pagoda2::Pagoda2$new) %>%
    createConos(sample.meta=sample.metadata, cell.meta=cell.metadata, k=40, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.1, alpha=0.1, font.size=c(2, 3))
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```

## MS

The threshold on transcripts here is set only because there was 1 almost empty cell reported, all other cells already had enough transcripts.

```{r, message=FALSE, warning=FALSE, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('MS/con.rds'), function() {
  mat <- Seurat::Read10X(DataPath("MS")) %>% .[rowSums(. > 0) >= 10,]
  meta <- read_delim(DataPath("MS/meta.txt"), delim='\t') %>%
    mutate(cell_type=gsub("-(Cntl|MS(-1|-2)?)", "", x=cell_type)) %>%
    rename(cellType=cell_type)

  sample.metadata <- meta[,5:14] %>% split(.$sample) %>% lapply(`[`, 1,) %>%
    do.call(rbind, .) %>% lapply(setNames, .$sample)
  cell.metadata <- meta %>% lapply(setNames, .$cell)

  mat.per.samp <- splitMatrixByFactor(mat, cell.metadata$sample)
  mat.per.cap <- splitMatrixByFactor(mat, cell.metadata$Capbatch)
  dub.info <- estimateDoubletInfo(mat.per.cap)
  p2s <- plapply(mat.per.samp, createPagoda, min.transcripts.per.cell=800, dub.scores=dub.info$scores,
                 dub.threshold=0.2, mc.preschedule=TRUE, n.cores=N_CORES, progress=FALSE)

  createConos(p2s, sample.meta=sample.metadata, cell.meta=cell.metadata, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.1, alpha=0.1)
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```

## PF

The paper already performed the filtration

```{r, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('PF/con.rds'), function() {
  cell.metadata <- read_csv(DataPath("PF/cell_metadata.csv")) %>% 
    dplyr::rename(cell=X1) %>% dplyr::filter(Diagnosis %in% c("Control", "IPF")) %>%
    dplyr::rename(sample=Sample_Name, cellType=celltype)

  sample.metadata <- cell.metadata %>%
    group_by(sample, Sample_Source, Diagnosis, Status, orig.ident) %>%
    summarise(n=n()) %>% dplyr::select(-n) %>%
    lapply(setNames, .$sample)

  mat <- DataPath("PF") %>% Seurat::Read10X(gene.column=1) %>%
    .[,cell.metadata$cell] %>% .[rowSums(. > 0) >= 10,]

  cell.metadata %<>% lapply(setNames, .$cell)
  mat.per.samp <- splitMatrixByFactor(mat, cell.metadata$sample)
  p2s <- plapply(mat.per.samp, createPagoda, mc.preschedule=TRUE, n.cores=N_CORES, progress=FALSE)
  createConos(p2s, sample.meta=sample.metadata, cell.meta=cell.metadata, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.1, alpha=0.1)
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```

## SCC

```{r, fig.width=5, fig.height=5}
con <- readOrCreate(DataPath('SCC/con.rds'), function() {
  mat <- data.table::fread(DataPath("SCC/counts.txt"), sep="\t") %>%
    {set_rownames(mltools::sparsify(.[3:nrow(.), 2:ncol(.)]), .$V1[3:nrow(.)])}

  cell.metadata <- read_delim(DataPath('SCC/cell_metadata.txt'), delim='\t') %>%
    filter(!(level3_celltype %in% c('Multiplet', 'Keratinocyte'))) %>%
    rename(cell=nCount_RNA, cellType=level3_celltype)
  cell.metadata$sample <- cell.metadata$cell %>% strsplit("_") %>%
    sapply(function(x) paste(x[1:2], collapse='_'))
  cell.metadata$cellType %<>% gsub("(Normal|Tumor)_", "", .)
  cell.metadata %<>% lapply(setNames, .$cell)

  mat <- mat[rowSums(mat > 0) >= 10, cell.metadata$cell]
  mat.per.samp <- splitMatrixByFactor(mat, cell.metadata$sample)
  mat.per.samp %<>% .[sapply(., ncol) > 500]

  p2s <- plapply(mat.per.samp, createPagoda, min.transcripts.per.cell=800,
                 mc.preschedule=TRUE, n.cores=N_CORES, progress=FALSE)

  createConos(p2s, sample.meta=NULL, cell.meta=cell.metadata, n.cores=N_CORES)
}, force=FORCE) %>% Conos$new()

con$plotGraph(groups=con$misc$cell_metadata$cellType, size=0.1, alpha=0.1)
```

```{r, message=FALSE, warning=FALSE}
rm(con); gc();
```
