---
title: 'Cacoa Report for the Multiple Sclerosis dataset'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, message=FALSE}
library(cacoa)
library(dataorganizer)

library(ggplot2)
library(dplyr)
library(cowplot)
library(Matrix)
library(dplyr)
library(magrittr)
library(readr)
library(org.Hs.eg.db)

theme_set(theme_bw())
```

```{r, message=FALSE, warning=FALSE}
cao <- read_rds(DataPath("MS/cao.rds")) %>% Cacoa$new()
cao$plot.params <- list(size=0.1, alpha=0.1, font.size=c(2, 3))
```

## Dependency on metadata

```{r, fig.width=8, fig.height=4}
lapply(c("coda", "expression.shifts"), function(sp) {
  cao$plotSampleDistances(space=sp, legend.position=c(1, 1))
}) %>% 
  plot_grid(plotlist=., ncol=2, labels=c("CoDA", "Expression"), hjust=0, label_x=0.02, label_y=0.99)
```

```{r, fig.width=7, fig.height=4}
sample_meta <- cao$data.object$misc$sample_metadata
lapply(c("coda", "expression.shifts"), function(sp) {
  smd <- as.data.frame(sample_meta) %>% dplyr::select(-sample, -diagnosis)
  sep.res <- cao$estimateMetadataSeparation(smd, space=sp, dist="l1", name=paste0("md.", sp),
                                            show.warning=FALSE)
  (-log10(sep.res$padjust)) %>% {tibble(Type=names(.), value=.)} %>%
    cacoa:::plotMeanMedValuesPerCellType(type="bar", yline=-log10(0.05), 
                                         ylab="-log10(separation p-value)") +
    scale_y_continuous(expand=c(0, 0, 0.05, 0)) +
    scale_fill_manual(values=rep("#2b8cbe", length(sample_meta))) +
    theme(axis.title.y=element_text(size=13))
}) %>% 
  plot_grid(plotlist=., ncol=2, labels=c("CoDA", "Expression"), hjust=0, label_x=0.2, label_y=0.98)
```

There is significant separation by Capbatch, Seqbatch and stage. Capbatch and Seqbatch are
very similar, so we show only the former.

```{r, fig.width=8, fig.height=8}
lapply(c('coda', 'expression'), function(sp) {
  lapply(c("Seqbatch", "stage"), function(n) {
    cao$plotSampleDistances(space=sp, legend.position=c(0, 1), sample.colors=sample_meta[[n]])
  }) %>% plot_grid(plotlist=., nrow=1)
}) %>% plot_grid(plotlist=., ncol=1)
```

CoDA space (top row) has even stronger separation by batch than expression space.

## Compositional differences

### Cluster-based

```{r}
cao$plotCellLoadings()
```

```{r}
cao$plotContrastTree()
```

We can see that most neuronal subtypes go to the left branch of the tree, while all glial 
and immune types go to the right.

### Cluster-free

```{r, fig.width=5, fig.height=5}
g0 <- cao$plotEmbedding(color.by='cell.groups')
g0
```

KDE-based:

```{r, fig.width=8, fig.height=4.5}
cao$plotCellDensity(name='cell.density.kde') %>% plot_grid(plotlist=., ncol=2)
```

Differential cell density:

```{r, fig.width=8, fig.height=4}
plot_grid(g0, cao$plotDiffCellDensity(name='cell.density.kde', legend.position=c(0, 1)), ncol=2)
```

Graph-based:

```{r, fig.width=8, fig.height=4.5}
cao$plotCellDensity(name='cell.density.graph') %>% plot_grid(plotlist=., ncol=2)
```

Differential cell density:

```{r, fig.width=8, fig.height=4}
plot_grid(g0, cao$plotDiffCellDensity(name='cell.density.graph', legend.position=c(0, 1)), ncol=2)
```

Here, graph-based method shows higher sensitivity, and its results mostly match to the 
cluster-based version. It also detects that in Astrocytes only one subtype is affected.

## Expression differences

### Cluster-based

All genes:

```{r}
cao$plotExpressionShiftMagnitudes()
```

No significant changes detected.

Top DE genes:

```{r}
cao$estimateExpressionShiftMagnitudes(n.permutations=5000, top.n.genes=500, n.pcs=8, 
                                      min.samp.per.type=4, name='es.top.de', verbose=FALSE)
```

```{r}
cao$plotExpressionShiftMagnitudes(name='es.top.de')
```

Number of DE genes:

```{r, message=FALSE, warning=FALSE}
cao$estimateDEPerCellType(
  independent.filtering=TRUE, test='DESeq2.Wald', verbose=FALSE, resampling.method='fix.samples', 
  fix.n.samples=6, n.cells.subsample=30, name='de.fix.samples', n.resamplings=50
)
```

```{r, warning=FALSE, message=FALSE}
cao$plotNumberOfDEGenes(
  name="de.fix.samples", type="box", show.resampling.results=TRUE, jitter.alpha=0.5,
  show.jitter=TRUE, y.offset=1
) + scale_y_log10(labels=c(0, 10, 100), breaks=c(1, 11, 101), expand=c(0, 0), limits=c(1, 300))
```

### Cluster-free

```{r, fig.width=8, fig.height=4}
cao$plotClusterFreeExpressionShifts(legend.position=c(0, 1), font.size=c(1, 2))
```

## Functional interpretation

```{r, fig.width=10, fig.height=12, warning=FALSE, message=FALSE}
cao$plotVolcano(xlim=c(-3, 3), ylim=c(0, 3.5), lf.cutoff=1)
```

```{r, fig.width=8, fig.height=8}
cao$plotOntologyHeatmapCollapsed(name="GSEA", genes="all", n=30, clust.method="ward.D")
```


```{r, message=FALSE, warning=FALSE}
cao$estimateGenePrograms(method="leiden", z.adj=TRUE, smooth=FALSE)
```

```{r, fig.width=12, fig.height=12}
cao$plotGeneProgramScores(
  legend.position=c(0, 1), plot.na=FALSE, 
  adj.list=theme(legend.key.width=unit(8, "pt"), legend.key.height=unit(12, "pt"))
)
```
