---
title: 'Simulation of pairwise distances'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, include=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(magrittr)
library(sccore)
library(cacoa)

theme_set(theme_bw())

devtools::load_all()

N_SAMPLES <- 10
N_CORES <- 10
```

In these simulations we artificially generate pairwise distance matrix by sampling from two
normal distributions with controlled distance between their means and controlled variance of
each distribution. Then, we compare three types of distance estimates:

1. **Shift**: $\textbf{d}_{between} - 0.5 \cdot \left( \mathrm{MED}(\textbf{d}_{case}) + \mathrm{MED}(\textbf{d}_{control}) \right)$
2. **Total**: $\textbf{d}_{between} - \mathrm{MED}(\textbf{d}_{control})$
3. **Variance**: $\textbf{d}_{case} - \mathrm{MED}(\textbf{d}_{control})$

## Equal variance, varying mean

First, let us change mean between two conditions, keeping the variance constant.

```{r, fig.width=6, fig.height=6, message=FALSE, warning=FALSE}
sim_df <- c(1, 2) %>% lapply(function(s) {
  plapply(seq(0.0, 2.0, 0.2), function(mo) {
    generateDistanceMatrixSimulations(mean.offset=mo, std.cnt=s, std.disease=s, n.samples=N_SAMPLES)
  }, n.cores=N_CORES, mc.preschedule=TRUE, fail.on.error=TRUE) %>% do.call(rbind, .)
}) %>% do.call(rbind, .)
sim_df$mean.offset %<>% paste0() %>% {factor(., levels=unique(.))}
sim_df$std.cnt %<>% paste0() %>% {factor(., levels=unique(.))}

plotSimulatedDistances(sim_df, "mean.offset", x.lab="Mean offset") +
  facet_wrap(~std.cnt, ncol=1, labeller=as_labeller(function(df) paste("Std =", df)))
```

The plot shows change in expression distances (y-axis) of different types (color) for 
increasing mean offset between case and control (x-axis) and fixed standard deviation 
(std=1.0 for the top plot and std=2.0 for the bottom plot). It can be seen that the "Variance" 
distance does depend on mean offset, while "Shift" and "Total" go up linearly. It is also
important to note that when variance within condition increase (bottom plot), the sensitivity
of all distances goes down. So, if two cell types have the same mean shift, 
but different variance, the cell type with lower variance would have higher normalized distance.

## Equal mean, varying variance

Now, we will fix the mean offset to different values and vary variance within Case samples,
keeping Control variance fixed (`std=1.0`).

```{r, fig.width=6, fig.height=9}
sim_df <- c(0, 1, 2) %>% lapply(function(m) {
  plapply(seq(1, 3.0, 0.2), function(std) {
    generateDistanceMatrixSimulations(mean.offset=m, std.disease=std, n.samples=N_SAMPLES)
  }, n.cores=N_CORES, mc.preschedule=TRUE, fail.on.error=TRUE) %>% do.call(rbind, .)
}) %>% do.call(rbind, .)
sim_df$mean.offset %<>% paste0() %>% {factor(., levels=unique(.))}
sim_df$std.disease %<>% paste0() %>% {factor(., levels=unique(.))}

plotSimulatedDistances(sim_df, "std.disease", "Case std") +
  facet_wrap(~mean.offset, ncol=1, labeller=as_labeller(function(df) paste("Mean offset =", df)))
```

This panel shows change in expression distances (y-axis) of different types (color) for 
increasing standard deviation in Case (x-axis) and fixed Mean offset (from 0 to 2 in the 
three plots top-down). It can be seen that the "Shift" distance is quite insensitive to 
the changes in variance, while "Variance" distance behaving exactly the same for each offset
value. It also shows that "Total" distance captures both changes.
