---
title: 'Simulation Note'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, include=FALSE, message=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(cacoa)
library(conos)
library(dataorganizer)
library(sccore)

devtools::load_all()

N_CORES <- 50
FORCE <- FALSE

DE_FRAC <- 0.1
LFC <- 1.5
N_CELLS <- 100
N_SAMPLES <- 8
N_REPEATS <- 30

MIN_SAMPS <- 3
N_PERMUTS <- 2500
N_PCS <- 6
TOP_N_GENES <- 500

theme_set(theme_bw() + theme(legend.background=element_blank()))
```

In these simulations we used muscat to generate artificial data from the Autism
dataset. It allowed us to vary individual covariates, fixing the amount of the
actual expression change. For each set of parameters we performed `r N_REPEATS`
repeats and estimated median distance and p-value (when available) for 
corresponding metrics.

```{r}
# Requires running simulation_types.Rmd first
sce <- read_rds(CachePath('asd_sim_sces.rds'))$`IN-PV`$prep
```

## Dependency on the number of cells and samples

### Number of samples per cell type

```{r, warning=FALSE, message=FALSE}
n_samples <- c(3, 5, 7, 9)
sims_samples <- generateSims(
  sce, n.cells=N_CELLS, de.frac=DE_FRAC, n.cores=N_CORES, lfc=LFC, 
  n.samples=n_samples, n.repeats=N_REPEATS, verbose=FALSE
)

cao_samples <- cacoaFromSim(sims_samples, n.cores=N_CORES)
cao_samples$estimateExpressionShiftMagnitudes(verbose=TRUE, n.permutations=N_PERMUTS, min.samp.per.type=MIN_SAMPS)
cao_samples$estimateExpressionShiftMagnitudes(
  verbose=TRUE, n.permutations=N_PERMUTS, top.n.genes=TOP_N_GENES, n.pcs=N_PCS, 
  min.samp.per.type=MIN_SAMPS, name='es.top.de'
)
cao_samples$estimateDEPerCellType(independent.filtering=TRUE, verbose=TRUE)
```

The raw expression distance does not depend on the number of samples:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df <- cao_samples$test.results$expression.shifts %>% 
  prepareExpressionShiftSimDf(sims=sims_samples) %>% mutate(ns=as.factor(ns)) 

plotExpressionShiftSimDf(p_df, x.col='ns', norm.dist=FALSE, adj.list=xlab("Num. samples"))
```

The normalized distance does not depend either, however the statisticsl power of 
the test depends a lot:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
plotExpressionShiftSimDf(p_df, x.col='ns', norm.dist=TRUE, adj.list=xlab("Num. samples"))
```

Estimating expression shifts over top DE introduce slight dependency though. Probably,
because of the quality of the selected DE genes.

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
cao_samples$test.results$es.top.de %>% 
  prepareExpressionShiftSimDf(sims=sims_samples) %>% mutate(ns=as.factor(ns)) %>% 
  plotExpressionShiftSimDf(x.col='ns', norm.dist=TRUE, adj.list=xlab("Num. samples"))
```

Cluster-free estimates are similar to cluster-based in this regard:

```{r}
cons_samples <- generateConsForClustFree(sims_samples, n.cores=N_CORES)
```

```{r}
cao_cf_per_type_samples <- cons_samples %$% 
  generateCacoaFromConsForClustFree(con.per.type, sim, n.cores=N_CORES)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
plotClusterFreeShiftSimulations(cao_cf_per_type_samples, params=cons_samples$sim$params, 
                                x.col='ns', adj.list=list(xlab('Num. samples')))
```

And the number of significant DE genes depends on the number of samples, just like
the significance tests for expression shifts above:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df$NumDE <- cao_samples$test.results$de %>% 
  sapply(function(de) sum(de$res$padj < 0.05)) %>% .[p_df$Type]

plotExpressionShiftSimDf(
  p_df, x.col='ns', dist.col='NumDE', adj.list=labs(x="Num. samples", y="Num. DE genes")
)
```

### Number of cells per cell type

```{r, message=FALSE, warning=FALSE}
n_cells <- c(25, 50, 75, 100, 125, 150)

sims_cells <- generateSims(
  sce, n.cells=n_cells, de.frac=DE_FRAC, n.cores=N_CORES, lfc=LFC, 
  n.samples=N_SAMPLES, n.repeats=N_REPEATS, verbose=FALSE
)

cao_cells <- cacoaFromSim(sims_cells, n.cores=N_CORES)
cao_cells$estimateExpressionShiftMagnitudes(verbose=FALSE, n.permutations=N_PERMUTS, min.samp.per.type=MIN_SAMPS)
cao_cells$estimateExpressionShiftMagnitudes(
  verbose=TRUE, n.permutations=N_PERMUTS, top.n.genes=TOP_N_GENES, n.pcs=N_PCS, 
  min.samp.per.type=MIN_SAMPS, name='es.top.de'
)
cao_cells$estimateDEPerCellType(independent.filtering=TRUE, verbose=TRUE)
```

The raw expression distances depend on the number of samples:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df <- cao_cells$test.results$expression.shifts %>% 
  prepareExpressionShiftSimDf(sims=sims_cells) %>% mutate(nc=as.factor(nc)) 

plotExpressionShiftSimDf(p_df, x.col='nc', norm.dist=FALSE, adj.list=xlab("Num. cells"))
```

However the normalization fixes it:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
plotExpressionShiftSimDf(p_df, x.col='nc', norm.dist=TRUE, adj.list=xlab("Num. cells"))
```

And here, using top-DE genes does not introduce new dependencies:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
cao_cells$test.results$es.top.de %>% 
  prepareExpressionShiftSimDf(sims=sims_cells) %>% mutate(nc=as.factor(nc)) %>% 
  plotExpressionShiftSimDf(x.col='nc', norm.dist=TRUE, adj.list=xlab("Num. cells"))
```

The same with cluster-free:

```{r, message=FALSE, warning=FALSE}
cons_cells <- generateConsForClustFree(sims_cells, n.cores=N_CORES, ncomps=15, metric="L2")
cao_cf_per_type_cells <- cons_cells %$% 
  generateCacoaFromConsForClustFree(con.per.type, sim, n.cores=N_CORES)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
plotClusterFreeShiftSimulations(cao_cf_per_type_cells, params=cons_cells$sim$params, 
                                x.col='nc', adj.list=list(xlab('Num. cells')))
```

While the number of significant DE genes does depend on the number of cells:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df$NumDE <- cao_cells$test.results$de %>% 
  sapply(function(de) sum(de$res$padj < 0.05)) %>% .[p_df$Type]

plotExpressionShiftSimDf(
  p_df, x.col='nc', dist.col='NumDE', adj.list=labs(x="Num. cells", y="Num. DE genes")
)
```
