---
title: 'Simulations for different cell types'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, include=FALSE, message=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(cacoa)
library(conos)
library(dataorganizer)
library(sccore)

devtools::load_all()

N_CORES <- 50
FORCE <- FALSE

DE_FRAC <- 0.1
LFC <- 1.5
N_CELLS <- 100
N_SAMPLES <- 8
N_REPEATS <- 30

MIN_SAMPS <- 3
N_PERMUTS <- 2500
N_PCS <- 6
TOP_N_GENES <- 500

theme_set(theme_bw() + theme(legend.background=element_blank()))
```

**WARNING:** this notebook takes hours of computations using 50 cores and requires
500+ Gb RAM.

In these simulations we used muscat to generate artificial data from the Autism
dataset. It allowed us to vary individual covariates, fixing the amount of the
actual expression change. For each set of parameters we performed `r N_REPEATS`
repeats and estimated median distance and p-value (when available) for
corresponding metrics.

## Prepare data

```{r}
con <- DataPath("ASD/con.rds") %>% readr::read_rds() %>% Conos$new()
cm <- con$getJointCountMatrix(raw=TRUE) %>% t()
sces <- readOrCreate(CachePath('asd_sim_sces.rds'), function() {
  sample.per.cell <- con$getDatasetPerCell()
  sample.groups <- con$misc$sample_metadata$diagnosis
  cell.groups <- con$misc$cell_metadata$cellType

  # Manually selected by coverage and to vary num. UMIs per type
  selected.types <- c("AST-FB", "Microglia", "Neu-NRGN", "OPC", "AST-PP",
                      "IN-SST", "IN-PV", "L2/3", "L5/6-CC")

  selected.types %>% setNames(., .) %>% mclapply(function(ct) {
    prepareSCEForSimulaton(cm, sample.per.cell=sample.per.cell, cell.groups=cell.groups,
                           sample.groups=sample.groups, ref.group="Control", subgroups=ct)
  }, mc.cores=N_CORES, mc.preschedule=TRUE)
}, force=FORCE)
```

## Dependency on the number of genes and other cell type covariates

Muscat doesn't allow varying number of genes per cell type properly, so we simulated
data from different cell types and visualized dependencies of the resulting distances
on the parameters of these cell types. It does not allow to analyze possible sources
of variation independently, but still provides some understanding of the dependencies.

```{r, warning=FALSE, message=FALSE}
n_cores_outer <- min(N_CORES, length(sces))
n_cores_inner <- max(N_CORES %/% length(sces), 1)

sims_types <- readOrCreate(CachePath('asd_sims_types.rds'), function() {
  lapply(c(0.0, 0.05, 0.1), function(def) {
    plapply(names(sces), function(n) {
      generateSims(sces[[n]]$prep, n.cells=N_CELLS, de.frac=def, n.cores=n_cores_inner, lfc=LFC,
                  n.samples=N_SAMPLES, suffix=n, n.repeats=N_REPEATS)
    }, n.cores=n_cores_outer, mc.allow.recursive=TRUE, progress=TRUE) %>% joinSims()
  })
}, force=FORCE)

caos_types <- readOrCreate(CachePath('asd_caos_types.rds'), function() {
  plapply(sims_types, function(sim) {
    cao <- suppressWarnings(cacoaFromSim(sim, n.cores=N_CORES %/% 3))
    cao$estimateExpressionShiftMagnitudes(verbose=FALSE, n.permutations=500, 
                                          min.samp.per.type=MIN_SAMPS)
    cao$estimateExpressionShiftMagnitudes(
      verbose=FALSE, n.permutations=500, top.n.genes=TOP_N_GENES, n.pcs=N_PCS,
      min.samp.per.type=MIN_SAMPS, name='es.top.de'
    )
    cao$estimateDEPerCellType(independent.filtering=TRUE, n.cores=N_CORES %/% 3, verbose=FALSE)
    cao
  }, n.cores=3, progress=TRUE, mc.allow.recursive=TRUE, mc.preschedule=TRUE)
}, force=FORCE)
```

```{r}
n_genes_per_type <- con$misc$cell_metadata$cellType %>% {split(names(.), .)} %>%
  sapply(function(ns) sum(colSums(cm[,ns]) > 0))

mean_expr_per_type <- con$misc$cell_metadata$cellType %>% {split(names(.), .)} %>%
  sapply(function(ns) sum(colSums(cm[,ns]))) %>% {. / n_genes_per_type[names(.)]}

covar_df <- mean_expr_per_type %>%
  {tibble(ME=., NG=n_genes_per_type[names(.)], Type=names(.))}

p_df <- lapply(1:3, function(i) {
  df <- prepareExpressionShiftSimDf(caos_types[[i]]$test.results$expression.shifts, 
                                    sims=sims_types[[i]])
  top.de.res <- caos_types[[i]]$test.results$es.top.de
  es.top <- top.de.res$dists.per.type %>% sapply(median)
  n.de <- sapply(caos_types[[i]]$test.results$de, function(de) sum(de$res$padj < 0.05))
  n.genes <- table(sims_types[[i]]$gene.info$cluster.id)
  df$NumDE <- n.de[df$Type]
  df$FracDE <- round(df$NumDE / n.genes[df$Type], 2)
  df$ESTop <- es.top[df$Type]
  df$ESTopPValue <- top.de.res$pvalues[df$Type]
  df
}) %>% bind_rows() %>% mutate(de.frac=as.factor(de.frac))

type_order <- p_df %>% filter(de.frac == '0') %>% group_by(suffix) %>%
  summarise(RawDist=median(RawDist)) %>% arrange(-RawDist) %>% .$suffix

p_df$suffix %<>% factor(levels=type_order)
```

```{r, fig.width=6, fig.height=4, warning=FALSE, message=FALSE}
plotExpressionShiftSimDf(p_df, x.col='suffix', norm.dist=FALSE, covar.col='de.frac',
                         covar.title='DE fraction') +
  cacoa:::theme_legend_position(c(1, 1)) + theme(axis.title.x=element_blank())
```

The plot shows raw expression distances between conditions (y-axis) simulated
from different cell types (x-axis) and different fraction of DE genes (color).
Each dot on the plot is median distance between the two conditions for one simulation.

It can be seen that there is a lot of variation for the same DE fraction, even if
it is set to 0.0. To explain the variation we may plot different cell type specific
covariates.

```{r, fig.width=7, fig.height=6, warning=FALSE, message=FALSE}
edr <- caos_types[[1]]$test.results$expression.shifts
tpg <- sims_types[[1]]$params %$% setNames(suffix, cluster.id)
p_df2 <- lapply(edr$p.dist.info, cacoa:::subsetDistanceMatrix,
                sample.groups=edr$sample.groups, cross.factor=FALSE) %>%
  sapply(median) %>% {tibble(Dist=., Cluster=names(.))} %>%
  mutate(Type=factor(tpg[Cluster], levels=type_order))

gg_within <- ggplot(p_df2, aes(x=Type, y=Dist)) +
  geom_boxplot(notch=TRUE, outlier.alpha=0) +
  ggbeeswarm::geom_quasirandom(size=0.1) +
  theme(panel.grid.major.x=element_blank(), axis.title.x=element_blank()) +
  ylab("Distances within condition")

ggs_covar <- mapply(function(yc, yl) {
  ggplot(na.omit(mutate(covar_df, Type=factor(Type, levels=type_order)))) +
    geom_bar(aes_string(x='Type', y=yc), stat="identity") +
    scale_y_continuous(expand=c(0, 0, 0.05, 0), name=yl) +
    theme(panel.grid.major.x=element_blank(), axis.title.x=element_blank())
}, c("ME", "NG"), c("Mean num. UMI", "Num. genes"), SIMPLIFY=FALSE)

plot_grid(gg_within, plotlist=ggs_covar, ncol=1, align='v')
```

This plot shows the number of expressed genes in real data (bottom), mean number
of molecules per gene in real data (middle) and median distance between samples
within the same condition for DE fraction set to 0.00 (top).

First, it can be seen that the distances within the same cell type are generally
higher for cell types with low coverage. Second, variation between samples within
the condition appear to be the main driver of the distances between the conditions.

Normalizing distances as we do to estimate expression shifts we can greatly reduce
those biases:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
p.theme <- theme(axis.text.x=element_text(angle=45, hjust=1), axis.title.x=element_blank()) +
    cacoa:::theme_legend_position(c(0, 1))
ggs <- plotExpressionShiftSimDf(p_df, x.col='suffix', norm.dist=TRUE, covar.col='de.frac',
                                covar.title='DE fraction', adj.list=list(p.theme), 
                                build.panel=FALSE)
ggs[[2]] %<>% {. + theme(legend.position="none")}
plot_grid(plotlist=ggs, ncol=2)
```

And here, focusing on top genes reduces biases even more:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
ggs <- p_df %>% mutate(NormDist=ESTop, pvalue=ESTopPValue) %>% 
  plotExpressionShiftSimDf(x.col='suffix', norm.dist=TRUE, covar.col='de.frac',
                           covar.title='DE fraction', adj.list=list(p.theme), 
                           build.panel=FALSE)
ggs[[2]] %<>% {. + theme(legend.position="none")}
plot_grid(plotlist=ggs, ncol=2)
```

```{r, message=FALSE, warning=FALSE}
cao_cf_per_type_types <- readOrCreate(CachePath('asd_cao_cf_per_type_types.rds'), function() {
  cons_types <- lapply(sims_types, generateConsForClustFree, n.cores=N_CORES)
  lapply(cons_types, function(ci) {
    generateCacoaFromConsForClustFree(ci$con.per.type, ci$sim, n.cores=N_CORES)
  })
}, force=FORCE)
```

Cluster-free estimates are the most sensitive to variation in the data. They still 
work to prioritize cell types with adequate variation, but are likely to miss variable
cell types:

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
cf_df <- lapply(1:3, function(pi) {
  caos <- cao_cf_per_type_types[[pi]]
  params <- sims_types[[pi]]$params
  z.scores <- sapply(caos, function(cao) 
    median(cao$test.results$cluster.free.expr.shifts$z_scores, na.rm=TRUE))
  sapply(caos, function(cao) 
      median(cao$test.results$cluster.free.expr.shifts$shifts, na.rm=TRUE)) %>%
    prepareExpressionDistDf(params=params) %>%
    inner_join(params, c("Type"="cluster.id")) %>%
    mutate(ns=factor(paste0(ns), levels=paste0(sort(unique(ns)))), z_score=z.scores[Type])
}) %>% bind_rows() %>% 
  mutate(de.frac=paste0(de.frac), suffix=factor(suffix, levels=levels(p_df$suffix)))

ggs <- mapply(function(dc, dn) {
  plotExpressionShiftSimDf(cf_df, x.col='suffix', dist.col=dc, covar.col='de.frac',
                           covar.title='DE fraction', adj.list=list(p.theme, ylab(dn)))
}, c('value', 'z_score'), c("Expression shift", "Z-score"), SIMPLIFY=FALSE)
ggs[[2]] %<>% + theme(legend.position="none")
plot_grid(plotlist=ggs, ncol=2)
```

Finally, below is the same plot for the number of significant DE genes. By design
of the simulations, the number of DE genes linearly depends on the total number
of genes. So, the right plot shows the fraction of DE genes, which also depends 
(slightly) on the variation within each cell types. It is up to debate, which of
these measures should be used for real world examples, though.

```{r, fig.width=10, fig.height=4, warning=FALSE, message=FALSE}
ggs <- mapply(function(yc, yl) {
  plotExpressionShiftSimDf(p_df, x.col='suffix', dist.col=yc, covar.col='de.frac',
                           covar.title='DE fraction') +
    scale_y_continuous(expand=c(0.01, 0, 0.05, 0), name=yl) +
    p.theme
}, c("NumDE", "FracDE"), c("Num. of DE genes", "Frac. of DE genes"), SIMPLIFY=FALSE)
ggs[[2]] %<>% + theme(legend.position="none")
plot_grid(plotlist=ggs, ncol=2)
```
