---
title: 'Simulations for sensitivity to LFC'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(cacoa)
library(conos)
library(dataorganizer)
library(sccore)

devtools::load_all()

N_CORES <- 50

DE_FRAC <- 0.05
LFC <- 1
N_CELLS <- 100
N_SAMPLES <- 8
N_REPEATS <- 30

MIN_SAMPS <- 3
N_PERMUTS <- 2500
N_PCS <- 6
TOP_N_GENES <- 500

theme_set(theme_bw() + theme(legend.background=element_blank()))
```

```{r}
# Requires running simulation_types.Rmd first
sce <- read_rds(CachePath('asd_sim_sces.rds'))$`IN-PV`$prep
```

## Sensitivity to LFC

Muscat does not allow simulating data with LFC < 1, so we analyse only values 
above 1. DE fraction is fixed to `r DE_FRAC`.

```{r, warning=FALSE, message=FALSE}
lfcs <- seq(1, 2, 0.25)
sims_lfc <- generateSims(
  sce, n.cells=N_CELLS, de.frac=DE_FRAC, n.cores=N_CORES, lfc=lfcs, 
  n.samples=N_SAMPLES, n.repeats=N_REPEATS, verbose=FALSE
)

cao_lfc <- cacoaFromSim(sims_lfc, n.cores=N_CORES)
cao_lfc$estimateExpressionShiftMagnitudes(verbose=TRUE, n.permutations=N_PERMUTS, 
                                          min.samp.per.type=MIN_SAMPS)
cao_lfc$estimateExpressionShiftMagnitudes(
  verbose=TRUE, n.permutations=N_PERMUTS, top.n.genes=TOP_N_GENES, n.pcs=N_PCS, 
  min.samp.per.type=MIN_SAMPS, name='es.top.de'
)
cao_lfc$estimateDEPerCellType(independent.filtering=TRUE, verbose=TRUE)
```


Increasing log2-fold change affects the distance a lot:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
p_df <- cao_lfc$test.results$expression.shifts %>% 
  prepareExpressionShiftSimDf(sims=sims_lfc) %>% mutate(lfc=as.factor(lfc))

plotExpressionShiftSimDf(p_df, x.col='lfc', norm.dist=TRUE, adj.list=xlab("Log2-fold change"))
```

Estimating expression shifts over top DE introduce slight dependency though. Probably,
because of the quality of the selected DE genes.

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
cao_lfc$test.results$es.top.de %>% 
  prepareExpressionShiftSimDf(sims=sims_lfc) %>% mutate(lfc=as.factor(lfc)) %>% 
  plotExpressionShiftSimDf(x.col='lfc', norm.dist=TRUE, adj.list=xlab("Log2-fold change"))
```

Cluster-free estimates have lower sensitivity, but still reach significance pretty 
fast:

```{r}
cao_cf_per_type_lfc <- generateConsForClustFree(sims_lfc, n.cores=N_CORES) %$% 
  generateCacoaFromConsForClustFree(con.per.type, sim, n.cores=N_CORES)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
plotClusterFreeShiftSimulations(cao_cf_per_type_lfc, params=sims_lfc$params, 
                                x.col='lfc', adj.list=list(xlab('Log2-fold change')))
```

The number of DE genes also gets more sensitive as the LFC increases:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df$NumDE <- cao_lfc$test.results$de %>% 
  sapply(function(de) sum(de$res$padj < 0.05)) %>% .[p_df$Type]

plotExpressionShiftSimDf(
  p_df, x.col='lfc', dist.col='NumDE', adj.list=labs(x="Log2-fold change", y="Num. DE genes")
)
```
