---
title: 'Simulations for sensitivity to DE fraction and LFC'
date: "`r Sys.Date()`"
author: "[Viktor Petukhov](https://github.com/VPetukhov/)"
output:
  html_notebook:
    toc: yes
    toc_depth: '3'
    toc_float: true
    df_print: paged
---

```{r libs, message=FALSE, warning=FALSE}
library(magrittr)
library(tidyverse)
library(cowplot)
library(cacoa)
library(conos)
library(dataorganizer)
library(sccore)

devtools::load_all()

N_CORES <- 50

DE_FRAC <- 0.05
LFC <- 1
N_CELLS <- 100
N_SAMPLES <- 8
N_REPEATS <- 30

MIN_SAMPS <- 3
N_PERMUTS <- 2500
N_PCS <- 6
TOP_N_GENES <- 500

theme_set(theme_bw() + theme(legend.background=element_blank()))
```

In these simulations we used muscat to generate artificial data from the Autism
dataset. It allowed us to vary individual covariates, fixing the amount of the
actual expression change. For each set of parameters we performed `r N_REPEATS`
repeats and estimated median distance and p-value (when available) for 
corresponding metrics.

```{r}
# Requires running simulation_types.Rmd first
sce <- read_rds(CachePath('asd_sim_sces.rds'))$`IN-PV`$prep
```

## Sensitivity to LFC and DE fraction

### DE fraction

```{r, message=FALSE, warning=FALSE}
de_fracs <- seq(0.0, 0.15, 0.025) %>% .[1:length(.)]

sims_fracs <- generateSims(
  sce, n.cells=N_CELLS, de.frac=de_fracs, n.cores=N_CORES, lfc=LFC, 
  n.samples=N_SAMPLES, n.repeats=N_REPEATS, verbose=FALSE
)

cao_fracs <- cacoaFromSim(sims_fracs, n.cores=N_CORES)
cao_fracs$estimateExpressionShiftMagnitudes(verbose=FALSE, n.permutations=N_PERMUTS, min.samp.per.type=MIN_SAMPS)
cao_fracs$estimateExpressionShiftMagnitudes(
  verbose=TRUE, n.permutations=N_PERMUTS, top.n.genes=TOP_N_GENES, n.pcs=N_PCS, 
  min.samp.per.type=MIN_SAMPS, name='es.top.de'
)
cao_fracs$estimateDEPerCellType(independent.filtering=TRUE, verbose=TRUE)
```


```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
p_df <- cao_fracs$test.results$expression.shifts %>% 
  prepareExpressionShiftSimDf(sims=sims_fracs) %>% mutate(de.frac=as.factor(de.frac)) 

plotExpressionShiftSimDf(p_df, x.col='de.frac', norm.dist=TRUE, adj.list=xlab("DE Fraction"))
```

After certain amount of changes, this distance reaches plateau, because adding
more DE genes does not contribute to the amount of changes in top. One can also
see that the sensitivity for small DE fraction improved greatly.

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
cao_fracs$test.results$es.top.de %>% 
  prepareExpressionShiftSimDf(sims=sims_fracs) %>% mutate(de.frac=as.factor(de.frac))  %>% 
  plotExpressionShiftSimDf(x.col='de.frac', norm.dist=TRUE, adj.list=xlab("DE Fraction"))
```

Cluster-free shifts are much less sensitive and more variable than the 
cluster-based version:

```{r, message=FALSE, warning=FALSE}
cao_cf_per_type_fracs <- generateConsForClustFree(sims_fracs, n.cores=N_CORES) %$%
  generateCacoaFromConsForClustFree(con.per.type, sim, n.cores=N_CORES)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
plotClusterFreeShiftSimulations(cao_cf_per_type_fracs, params=sims_fracs$params, 
                                x.col='de.frac', adj.list=list(xlab('DE fraction')))
```

There is no surprises regarding the number of DE genes as a metric here:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df$NumDE <- cao_fracs$test.results$de %>% 
  sapply(function(de) sum(de$res$padj < 0.05)) %>% .[p_df$Type]

plotExpressionShiftSimDf(
  p_df, x.col='de.frac', dist.col='NumDE', adj.list=labs(x="DE fraction", y="Num. DE genes")
)
```

```{r, message=FALSE, warning=FALSE}
rm(cao_cf_per_type_fracs, cao_fracs)
gc()
```

### Log2-fold change (LFC)

Muscat does not allow simulating data with LFC < 1, so we analyse only values 
above 1. DE fraction is fixed to `r DE_FRAC`.

```{r, warning=FALSE, message=FALSE}
lfcs <- seq(1, 2, 0.25)
sims_lfc <- generateSims(
  sce, n.cells=N_CELLS, de.frac=DE_FRAC, n.cores=N_CORES, lfc=lfcs, 
  n.samples=N_SAMPLES, n.repeats=N_REPEATS, verbose=FALSE
)

cao_lfc <- cacoaFromSim(sims_lfc, n.cores=N_CORES)
cao_lfc$estimateExpressionShiftMagnitudes(verbose=TRUE, n.permutations=N_PERMUTS, 
                                          min.samp.per.type=MIN_SAMPS)
cao_lfc$estimateExpressionShiftMagnitudes(
  verbose=TRUE, n.permutations=N_PERMUTS, top.n.genes=TOP_N_GENES, n.pcs=N_PCS, 
  min.samp.per.type=MIN_SAMPS, name='es.top.de'
)
cao_lfc$estimateDEPerCellType(independent.filtering=TRUE, verbose=TRUE)
```


Increasing log2-fold change affects the distance a lot:

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
p_df <- cao_lfc$test.results$expression.shifts %>% 
  prepareExpressionShiftSimDf(sims=sims_lfc) %>% mutate(lfc=as.factor(lfc))

plotExpressionShiftSimDf(p_df, x.col='lfc', norm.dist=TRUE, adj.list=xlab("Log2-fold change"))
```

Estimating expression shifts over top DE introduce slight dependency though. Probably,
because of the quality of the selected DE genes.

```{r, fig.width=9, fig.height=4, warning=FALSE, message=FALSE}
cao_lfc$test.results$es.top.de %>% 
  prepareExpressionShiftSimDf(sims=sims_lfc) %>% mutate(lfc=as.factor(lfc)) %>% 
  plotExpressionShiftSimDf(x.col='lfc', norm.dist=TRUE, adj.list=xlab("Log2-fold change"))
```

Cluster-free estimates have lower sensitivity, but still reach significance pretty 
fast:

```{r}
cao_cf_per_type_lfc <- generateConsForClustFree(sims_lfc, n.cores=N_CORES) %$% 
  generateCacoaFromConsForClustFree(con.per.type, sim, n.cores=N_CORES)
```

```{r, message=FALSE, warning=FALSE, fig.width=8, fig.height=3}
plotClusterFreeShiftSimulations(cao_cf_per_type_lfc, params=sims_lfc$params, 
                                x.col='lfc', adj.list=list(xlab('Log2-fold change')))
```

The number of DE genes also gets more sensitive as the LFC increases:

```{r, fig.width=5, fig.height=4, warning=FALSE, message=FALSE}
p_df$NumDE <- cao_lfc$test.results$de %>% 
  sapply(function(de) sum(de$res$padj < 0.05)) %>% .[p_df$Type]

plotExpressionShiftSimDf(
  p_df, x.col='lfc', dist.col='NumDE', adj.list=labs(x="Log2-fold change", y="Num. DE genes")
)
```
